---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```

```{r badges, results='asis', echo=FALSE}
suppressPackageStartupMessages(library(badgecreatr))
acc <- "MarselScheer"
repo <- "simTool"
branch <- "master"
status <- "active"
lifecycle <- "Stable"
badges <- list(
  badge_travis(ghaccount = acc, ghrepo = repo, branch = branch),
  badge_lifecycle(lifecycle = lifecycle),
  badge_projectstatus(status = status),
  badge_cran_version_ago(packagename = repo),
  badge_cran_downloads(packagename = repo),
  badge_license()
  )
cat(paste(badges, collapse = " "))
```

# simTool

An R-Package that facilitates simulation studies. It disengages the researcher from administrative source code.

The *simTool* package is designed for statistical simulations that
have two components. One component generates the data and the other one
analyzes the data. The main aims of the *simTool* package are the reduction
of the administrative source code (mainly loops and management code for the results) 
and a simple applicability
of the package that allows the user to quickly learn how to work with the
*simTool* package. Parallel computing is also supported. Finally, convenient
functions are provided to summarize the simulation results.

## Example

This small simulation (using 4 cores) illustrates how the confidence interval based on the
t-distribution performs on exponential distributed random variables. The
following lines generate exponential distributed random variables of size
10, 50, 100, and 1000. Afterwards the *t.test* using confidence levels
0.8, 0.9, 0.95 are applied. This is repeated 1000 times to estimate
the coverage:

```{r example}
library(simTool)
dg <- expand_tibble(fun = "rexp", rate = 10, n = c(10L, 50L, 100L, 1000L))
pg <- expand_tibble(proc = "t.test", conf.level = c(0.8, 0.9, 0.95))
et <- eval_tibbles(dg, pg, 
  ncpus = 4,
  replications = 10^3,
  post_analyze = function(ttest) tibble::tibble(
    coverage = ttest$conf.int[1] <= 1 / 10 && 1 / 10 <= ttest$conf.int[2]),
  summary_fun = list(mean = mean)
)
et
```

## Installation

You can install simTool from github with:

```{r gh-installation, eval = FALSE}
install.packages("simTool")
# install.packages("devtools")
devtools::install_github("MarselScheer/simTool")
```

Or from CRAN with:

```{r cran-installation, eval = FALSE}
install.packages("simTool")
```

## Test suite

With tinytest you can rerun the test for simTool by calling

```{r tinytest, eval = FALSE}
# install.packages("tinytest")
tinytest::test_package("simTool")
```
