---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```


[![R build status](https://github.com/MarselScheer/simTool/workflows/R-CMD-check/badge.svg)](https://github.com/MarselScheer/simTool/actions)
[![codecov](https://codecov.io/gh/MarselScheer/simTool/branch/development/graph/badge.svg?token=aCoLwwqLrB)](https://app.codecov.io/gh/MarselScheer/simTool/)
[![lifecycle](https://img.shields.io/badge/lifecycle-stable-brightgreen.svg)](https://lifecycle.r-lib.org/articles/stages.html)
[![Project Status: Active – The project has reached a stable, usable
state and is being actively
developed.](https://www.repostatus.org/badges/latest/active.svg)](https://www.repostatus.org/#active)
[![CRAN_Status_Badge](https://www.r-pkg.org/badges/version/simTool)](https://cran.r-project.org/package=simTool)
[![metacran
downloads](https://cranlogs.r-pkg.org/badges/simTool)](https://cran.r-project.org/package=simTool)
[![license](https://img.shields.io/badge/license-GPL--3-blue.svg)](https://www.gnu.org/licenses/gpl-3.0.en.html)


# simTool

An R-Package that facilitates simulation studies. It disengages the researcher from administrative source code.

The *simTool* package is designed for statistical simulations that
have two components. One component generates the data and the other one
analyzes the data. The main aims of the *simTool* package are the reduction
of the administrative source code (mainly loops and management code for the results) 
and a simple applicability
of the package that allows the user to quickly learn how to work with the
*simTool* package. Parallel computing is also supported. Finally, convenient
functions are provided to summarize the simulation results.

## Example

This small simulation (using 4 cores) illustrates how the confidence interval based on the
t-distribution performs on exponential distributed random variables. The
following lines generate exponential distributed random variables of size
10, 50, 100, and 1000. Afterwards the *t.test* using confidence levels
0.8, 0.9, 0.95 are applied. This is repeated 1000 times to estimate
the coverage:

```{r example}
library(simTool)
dg <- expand_tibble(fun = "rexp", rate = 10, n = c(10L, 50L, 100L, 1000L))
pg <- expand_tibble(proc = "t.test", conf.level = c(0.8, 0.9, 0.95))
et <- eval_tibbles(dg, pg, 
  ncpus = 4,
  replications = 10^3,
  post_analyze = function(ttest) tibble::tibble(
    coverage = ttest$conf.int[1] <= 1 / 10 && 1 / 10 <= ttest$conf.int[2]),
  summary_fun = list(mean = mean)
)
et
```

## Installation

You can install simTool from github with:

```{r gh-installation, eval = FALSE}
remotes::install_github("MarselScheer/simTool")
```

Or from CRAN with:

```{r cran-installation, eval = FALSE}
install.packages("simTool")
```

## Test suite

With tinytest you can rerun the test for simTool by calling

```{r tinytest, eval = FALSE}
# install.packages("tinytest")
tinytest::test_package("simTool")
```
